Seismic parameter estimation and the Canadian crust

### 2013-09-22 paper/draftBP.tex
discussion draft

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{discontinuities}
  \caption{}
  \label{fig:disc}
\end{figure}

@@

Seismic reflectivity of the continental crust is a valuable constraint on the processes governing crustal evolution. Early efforts to characterize continental crust revolved around a slab model with a lower velocity upper crust and a higher velocity lower crust [Conrad, 1925; Jeffreys, 1926]. Evidence for this model was originally put forth by Conrad [1925] who suggested observable P arrivals were propogating along a mid-crustal interface similar to the deeper crust-mantle interface imaged by Andrija Mohorovičić [Litak and Brown, 1989]. Both the crustal slab model and the Conrad Discontinuity, as the mid-crustal interface became known, persisted in the literature but were increasingly criticized as being too simplistic, in the case of the slab model, or, for the Conrad Discontinuity, lacking consistent evidence [Litak and Brown, 1989]. The development of reflection seismology allowed workers to further argue for a complex heterogenous crust and propose a model whereby smooth random lateral inhomogenities superimposed on a linearly increasing velocity gradient could produce seismic responses similar to that of a simple two layer model [Litak and Brown, 1989; Mereu and Ojo, 1981].

Despite the fading support for the Conrad Discontinuity as a discrete geological phenomena the generalization of the crust into seismically distinguishable upper and lower components continued. Mooney and Brocher [1987] argued for a model of the continental crust seperated into a seismically reflective lower crust overlain by transparent upper crust. Their physical interpretion is of a poorly laminated upper crust dominated by features of short length scale contrasted with a lower crust composed of thin (100 metre) lamellae of significant length scales. Supporting evidence for a reflective lower crust [Ito, 1998; Mayer, 1996; Meissner et. al., 2006] indicate that this is likely regional and representational of extented as well as thicker warmer crusts.

@@

The lower energy noted in Section \ref{section:discontinuities} in the 20 to 30 km depth window for the station profiles given in figure \ref{fig:disc} provides some evidence for seismically distinguishable crustal layers. To investigate this further we stretch and interpolate the profiles so that the Moho of each profile lines up on a common axis (Figure \ref{fig:discInterp}, left) and then take the magnitude of each profile and stack (Figure \ref{fig:discInterp}, right). The energy of the stack is most pronounced near the surface, associated with sedimentary layers, and at the Moho. There is a marked decrease in energy in the bottom third of the profile.

(Jackson 2002 - strength of lithosphere - mirrors decrease in energy)

2. cite station sorted figure + perhaps normalized stack figure (as per above) and discuss where
energy lies, most pronounced near surface and Moho, lower energy levels in bottom 1/3

3. draw association between bottom 1/3 of low teleseismic reflectivity and high lower crustal  (seismic
reflection) reflectivity, explain effect of frequency in different response

4. discuss evidence for Conrad as distinct discontinuity (doesn't appear to be) versus the more
general subdivision of upper and lower crusts with distinct properties.

Mean crustal composition is an important parameter characterizing continental crust that should not be confused with actual lithological or mineralogical state of the crust. Rather, mean crustal composition is a representative average quantity that serves as a constraint on crustal structure and evolution in space and time. Early attempts at estimating mean crustal composition were largely performed through extrapolations based on the surface sampling of upper crustal rocks [Rudnick and Gao, 2003]. After the plate-tectonic revolution, Taylor [1967, 1977], proposed the 'island arc' or 'andesite' model which posited that bulk crustal composition is equivalent to the average convergent-margin andesite. Several problems with this theory subsequently emerged; for example, average surficial island arc composition is more basaltic than andesitic and any crustal growth via basaltic basal accretion will only serve to magnify this compositional contrast. Hence this model cannot account for the intermediate composition of post-Archean crust. Recognition of these shortcomings led to the analysis of data from crustal refraction seismic surveys to interrogate middle and lower crustal levels and place constraints on their composition [Christensen and Mooney, 1995; Rudnick and Fountain, 1995]. Unlike these latter studies, the present work adopts a bulk crustal seismic analysis along the lines of Smithson et. al. [1981] that combines $V_P/V_S$ from teleseismic measurements with contraints on $V_P$ from previous active source studies as presented in the crust 1.0 model.

>>>>

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{discontinuity}
  \caption{}
  \label{fig:disc}
\end{figure}

@@

Seismic reflectivity of the continental crust is a valuable constraint on the processes governing crustal evolution. Early efforts to characterize continental crust revolved around a slab model with a lower velocity upper crust and a higher velocity lower crust [Conrad, 1925; Jeffreys, 1926]. Evidence for this model was originally put forth by Conrad [1925] who suggested observable P arrivals were propagating along a mid-crustal interface similar to the deeper crust-mantle interface imaged by Andrija Mohorovičić [Litak and Brown, 1989]. Both the crustal slab model and the Conrad Discontinuity, as the mid-crustal interface became known, persisted in the literature but were increasingly criticized as being too simplistic, in the case of the slab model, or, for the Conrad Discontinuity, lacking consistent evidence [Litak and Brown, 1989]. The development of reflection seismology allowed workers to further argue for a complex heterogeneous crust and propose a model whereby smooth random lateral inhomogeneities superimposed on a linearly increasing velocity gradient could produce seismic responses similar to that of a simple two layer model [Litak and Brown, 1989; Mereu and Ojo, 1981].

Despite the fading support for the Conrad Discontinuity as a discrete geological phenomena the generalization of the crust into seismically distinguishable upper and lower components continued. Mooney and Brocher [1987] argued for a model of the continental crust separated into a seismically reflective lower crust overlain by transparent upper crust. Their physical interpretation is of a poorly laminated upper crust dominated by features of short length scale contrasted with a lower crust composed of thin (100 metre) lamellae of significant length scales. Supporting evidence for a reflective lower crust confirms widespread laminated structure in the lower crust with thicknesses of 100m to 300m though these structures have been found to be localized and representational of extended as well as thicker warmer crusts [Ito, 1998; Mayer, 1996; Meissner et. al., 2006].

@@

The lower energy noted in Section \ref{section:discontinuities} in the 20km to 30km depth window for the station profiles given in figure \ref{fig:disc} provides some evidence for seismically distinguishable crustal layers. To investigate this further we stretch and interpolate the profiles so that the Moho of each profile lines up on a common axis (Figure \ref{fig:discInterp}, left) and then take the magnitude of each profile and stack (Figure \ref{fig:discInterp}, right). The energy of the stack is most pronounced near the surface, associated with sedimentary layers, and at the Moho. There is a marked decrease in energy in the bottom third of the profile. This study employs S-wave energy using both direct and reflected arrivals and it has been shown that these waves interact with velocity gradients on scales lengths in the order of $\lambda / 2$ for direct phases and $\lambda / 4$ for reflected phases where $\lambda$ is wavelength [Bostock, 1999]. For 1Hz teleseismic S-waves crustal structure in the scale lengths of 750m to 1500m is resolvable. Higher frequency active source data (10Hz - 40Hz) employing P-wave sources would be sensitive to structure roughly in the ranges of 75m to 300m. Lithoprobe transects [Hammer et. al., 2010] show significant heterogeneity throughout the crust including several transects, such as the Western Superior transect, that appear to contain higher reflectivity in the lower crust. One interpretation for these data, taken together, is that in general the Canadian lower crustal structure contains fewer thick (~1km) velocity contrasts as compared to the upper crust. This interpretation allows for high reflectivity in the lower crust arising from thinner, less than 750m, contrasts.

Though the Conrad Discontinuity continues to be invoked, for example in a study of the Rhine Graben [Mayer, 1996], as terminology to describe the interface between a transparent upper crust and reflective lower crust, it does not appear to be a prevalent feature of the Canadian crust. The results from this study do suggest however, that in general, and with the seismic methods employed here, a distinction between the seismic properties of the upper crust and the bottom third of the crust can be resolved. Studies on crustal strength [Jackson, 2002] and on brittle and ductile models [Burov and Watts, 2006; Meisner and Rabbel, 1999] as well as the widely noted reduced seismicity of the lower crust indicate changes in crustal processes at similar depths, though see [Deichmann, 1991; Handy and Brun, 2004; Simpson, 1999] for competing evidence.

Mean crustal composition is another important parameter characterizing continental crust though it is important to distinguish it from actual lithological or mineralogical state of the crust. Rather, mean crustal composition is a representative average quantity that serves as a constraint on crustal structure and evolution in space and time. Early attempts at estimating mean crustal composition were largely performed through extrapolations based on the surface sampling of upper crustal rocks [Rudnick and Gao, 2003]. After the plate-tectonic revolution, Taylor [1967, 1977], proposed the 'island arc' or 'andesite' model which posited that bulk crustal composition is equivalent to the average convergent-margin andesite. Several problems with this theory subsequently emerged; for example, average surficial island arc composition is more basaltic than andesitic and any crustal growth via basaltic basal accretion will only serve to magnify this compositional contrast. Hence this model cannot account for the intermediate composition of post-Archean crust. Recognition of these shortcomings led to the analysis of data from crustal refraction seismic surveys to interrogate middle and lower crustal levels and place constraints on their composition [Christensen and Mooney, 1995; Rudnick and Fountain, 1995]. Unlike these latter studies, the present work adopts a bulk crustal seismic analysis along the lines of Smithson et. al. [1981] that combines $V_P/V_S$ from teleseismic measurements with constraints on $V_P$ from previous active source studies as presented in the crust 1.0 model.
